\documentclass[11pt]{article}
\usepackage{fullpage}

\bibliographystyle{plain}

\input{WebMacros}
\input{SMacros}
\input{CMacros}
\input{XMLMacros}

\begin{document}

\title{GraphLayout: graph layout as a plugin in {\tt ggobi}}
\author{Deborah F. Swayne}
\maketitle

\begin{abstract}
GraphLayout is a plugin for ggobi, and this manual
describes its use.  It can be expected to change and grow.
\end{abstract}

\section{The data}

First, you need some data.  The ascii format supported by ggobi is not
adequate for the specification of edges, so you'll probably use an xml
file with a minimum of two datasets:  a set of cases or nodes, and a set
of edges.  The records in the node data must have
\XMLAttribute{id}s:

\begin{verbatim}
<record id="0"> ... </record>
<record id="1"> ... </record>
\end{verbatim}

The edge records use those \XMLAttribute{id}s to specify a
\XMLAttribute{source} and \XMLAttribute{destination}:

\begin{verbatim}
<record source="0" destination="1"> ... </record>
\end{verbatim}

No two records in the same xml file may have the same record id.

The sample data that is discussed here is part of the ggobi distribution:
\file{snetwork.xml}, an artificial social network of 140 people who are
connected by 205 edges, representing some (unspecified) form of social
contact.  Notice that there are two variables recorded for each node,
and two variables for each edge.

In a ggobi scatterplot display, use the {\bf Edges} menu to
display the edges, and maybe the ``arrowheads'' which indicate
edge direction.  Note: Until the edges are displayed, neither layout
algorithm will work.

\section{Layout}

There are three layout algorithms presently implemented in GraphLayout.
A radial layout method is available without the use of any external
libraries, and two more methods (called dot and neato) are available
if you build GraphLayout with graphviz (www.graphviz.org).

Find the ``Graph layout ...'' item on ggobi's Tools menu
to open the GraphLayout control panel.  It contains a ``notebook''
widget with one page for each layout method.

Each layout method generates a new dataset when applied, and opens
a new scatterplot displaying a plot of the node positions.  The
scatterplots are linked across both nodes and edges.

\subsection {Radial layout}

The radial layout \cite{Wills99} sets the first node in the dataset
as the center node, and arranges the rest of the nodes in concentric
circles around it.  The resulting layout is a tree arranged radially.
If the underlying graph is not very tree-like, the layout can result
in a great many edge crossings, and the layout doesn't do anything to
minimize these crossings.

Click on ``apply'' under the ``Radial'' tab.  It generates a new
dataset with the following variables:

\begin{itemize}
\item x, y: The node positions.  Plot y against x to view the layout.
\item depth: For each node, the number of steps required to get from it
  to the center node. (Plots of y or x against depth unwrap the
  radial layout and make it look like a tree, if an unbeautiful one.)
\item in degree, out degree:  The number of edges in and out of each
  node. (This is the only place edge direction shows up.)
\item nparents: For $node_i$, the number of nodes with $depth = depth_i~-~1$
  with which it shares an edge. 
\item nchildren: For $node_i$, the number of nodes with $depth = depth_i~+~1$
  with which it shares an edge.
\item nsiblings: For $node_i$, the number of nodes with $depth = depth_i$
  with which it shares an edge.
\end{itemize}

\subsection {Graphviz layouts}

There are two layouts available with the graphviz {\cite
GansnerNorth00} package:  {\bf dot} makes hierarchical layouts of
directed graphs, and {\bf neato} makes ``spring'' model layouts of
undirected graphs.   The software can be obtained at www.graphviz.org,
and some instructions for building it for ggobi are in the INSTALL
file in the GraphLayout source directory.

When you have built GraphLayout with the graphviz package, the plugin
window will include a tab labelled ``graphviz,'' with a button for
each of the two layout methods.  In each case, a new dataset with two
variables is generated (the horizontal and vertical coordinates of
the node positions).

\section{Manipulations}

All the standard ggobi direct manipulations are available.  Plots of
the graph can obviously be linked node-wise to plots of node covariates.
What might be less obvious is that they can also be linked to plots of
edge covariates, so that an edge in the graph corresponds to a node in the
plot of edge data.  Nodes can be interactively brushed and ``identified;''
edges can be brushed -- to set the color of the line type and width.
The ``color schemes'' tool can be used to automatically color the nodes
or the edges.

Nodes can also be moved, which can help untangle the layout a bit.
Groups of nodes can be moved together: first brush them with
the same symbol and color, and then select {\tt Move brush group}
in the {\tt Move points} ViewMode.

There is only one added manipulation, and that puts it an appearance
during the use of identification.  When a node is made ``sticky,'' a
set of nodes and edges are are highlighted using the current brushing
color:  the sticky node, its edges, and all nodes to which those edges
are connected.  To remove the highlighting, click again on the node to
make it unsticky.

\subsection{Thinning the graph}

Brushing can be used to thin the plot, by hiding nodes with especially
high in or out degree, for example, or nodes with large values of depth.
Once those nodes are hidden, a new radial layout can be produced.
Any nodes or edges still visible that no longer have a path to the
center node will then be hidden as well, so it isn't necessary to erase
an entire subtree: erasing its root is enough.

If you erase the center node, GraphLayout will refuse to proceed with
a new layout until you have restored it.

\section{Plot control from R}

If you have launched ggobi from within R, you can use the API to
drive the plot.  Here are some fragments of code in the S language
that I have used.

In the first example, I have two xml files representing two
related graphs, and I'm interested in comparing them.  This is
part of the code used to highlight the edges the two graphs
have in common.

\begin{verbatim}
g1 <- ggobi("f1.xml")
setDisplayEdges.ggobi(.gobi=g1)
e1 <- getEdges.ggobi(.data=2, .gobi=g1)
g2 <- ggobi("f2.xml")
setDisplayEdges.ggobi(.gobi=g2)
e2 <- getEdges.ggobi(.data=2, .gobi=g2)
...
esame1 <- enames1 %in% enames2
edgecolors1 <- rep(1, nedges1)
edgecolors1[esame1==T] <- 6
setColors.ggobi (edgecolors1, .data=2, .gobi=g1)
...
\end{verbatim}


In the second example, there is one graph, and its edge covariates are
the values of a variable recorded for each edge at $t_i$.  I use R to
animate the graph, using color to encode the edge weight; I first chose
a sequential colorscheme.  Similarly, one could use setGlyphs.ggobi() to
set node type or size.  The same command sets edge type or thickness
when applied to the edge data.  To hide some of the edges, use
setHiddenCases.ggobi().

\begin{verbatim}
edges <- getData.ggobi(2)
ntimesteps <- dim(edges)[2]

for (i in 1:ntimesteps) {
  tgraph (i, edges)
  ...
  colors <- integer(dim(edges)[1])
  colors[lnw>(3*mx/4)] <- 5
  colors[lnw>(mx/2) & lnw<=(3*mx/4)] <- 4
  colors[lnw>(mx/4) & lnw<=(mx/2)] <- 3
  colors[lnw>0 & lnw<=(mx/4)] <- 2
  setColors.ggobi (colors, 1:length(colors), 2)
}
\end{verbatim}

In an extension of the second example, the xml file includes three
datasets.  The third is of dimension $ntimesteps$ by $2$, and its
single time series plot represents the sum of all measurements
for all edges at each time step.  As the animation runs,
we highlight the corresponding point in a scatterplot of this time series.

\begin{verbatim}
tcolors <- integer(ntimesteps)
tcolors[1:length(tcolors)] <- 3
tcolors[i] <- 7
setColors.ggobi (tcolors, 1:length(tcolors), 3)
\end{verbatim}

\section{Future work}

We want very much to expand the set of available layout methods.
We intend to port {\tt xgvis}, an extension to {\tt ggobi's} predecessor,
{\tt xgobi,} a rich environment for interactive multidimensional scaling.

% Bibliography: MDS reference, Graham Wills layout paper, xgvis papers

\bibliography{references}

\end{document}
